---
title: "HW2-EDA with R - Part1: Plotting"
author: "misken"
date: "January 25, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.

## Step 3 - Set working directory and explore data folder

Set the working directory or create an R Studio project in the current folder. You'll notice that there is a folder named **data**.
Inside of it you'll find the data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data. **You do NOT need to rerun the census_data_prep.R file.** I'm
including it and encouraging you to take a look through it to get a sense of what a
typical data prep script might look like. We'll be learning to do all of the things
shown in that file.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.

Note that if you have errors in your R code, **knitr** will stop at the error
and won't create an html file. Sooo, you should comment out R code that has
errors and write a little comment stating that it's an error. Keep doing this
until you successfully knit an html file based on your R Markdown document.
You'll be submitting both the html file as well as the Rmd file.

You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(dplyr)
library(ggplot2)
```


### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_MI.csv** into a data frame named county_MI.

```{r read_countydata}
county_df <- read.csv("data/county.csv", header = TRUE)
county_MI <- read.csv("data/county_MI.csv", header = TRUE)
```

* What does each row represent?
Each row represents a different county's (observation) census results.

Check the structure of each data.frame and use the `summary` function to begin
exploring the data. 

* Are there any NA values? 
county_df has NA values for Income, IncomeErr, ChildPoverty.
* What differences do you notice between the structure of the two data frames?
county_MI has a variable for Peninsula, whereas county_df does not.The county_df starts with state, whereas county_MI starts with county (the first columns are in different orders)

```{r}
str(county_df)
str(county_MI)
summary(county_df)
summary(county_MI)
```

The county_MI data frame contains data for Michigan only and includes an
peninsula field. See **census_data_prep.R** to see how this field was
created. It involves scraping a table out of a web page.

**Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.**

### Problem 2 - Basic plots for median Income

Note that the field called `Income` is the median income
for that county. You do **NOT** need to compute a median. The values are the medians.

Create a histogram, boxplot, kernel density plot and violin plot for median
income for the entire US. 
Write a short paragraph describing the distribution of median incomes.
The distribution is right-skewed with the majority of data falling between 30000-55000. Outliers reach all the way to 125000. The mode is near 42000. From the boxplot, the IQR is just over 12500 wide, with the median of Median Incomes hovering near 42000 (however, this NOT the median of all households).

```{r}
ggplot(county_df) + geom_histogram(aes(x = Income)) + xlab("Median Income") +ggtitle("Median Income Distribution") + ylab("Median Income")
ggplot(county_df) + geom_boxplot(aes(x = "", y = Income)) + xlab("") + ylab("Median Income")+ggtitle("Median Income Distribution") + ylab("Median Income")
ggplot(county_df) + geom_density(aes(x = Income))+ggtitle("Median Income Distribution") + ylab("Median Income")
ggplot(county_df) + geom_violin(aes(x = "", y = Income)) + xlab("")+ggtitle("Median Income Distribution") + ylab("Median Income")
```

### Problem 3 - Histogram of total population

Create a histogram of total population for counties in the US. 

```{r histo_totpop}
ggplot(county_df) + geom_histogram(aes(x = TotalPop)) +xlab("Total Population") + ggtitle("Histogram of Total Population")
```

Repeat the histogram about but make the following style changes to the plot.

* Add a title such as "Distribution of population for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Total Population"
* Change the colors of the bars to something other than the default black

```{r histo_stylechanges}
ggplot(county_df) + geom_histogram(aes(x = TotalPop), fill = "turquoise") + ggtitle("Distribution of population for US counties") + ylab("Number of counties") + xlab("Total Population")
```

Repeat the histogram but only for counties with less than 1.5 million people. Do NOT use
dplyr to do the filtering - instead, use boolean indexing (see intro to dataframes Rmd file).

```{r histo_lt1.5m}
#test <- county_df$TotalPop < 1500000
#test
# county_df %>%
#   subset(TotalPop < 1500000) %>%
# ggplot(aes(x = county_df$TotalPop), fill = "turquoise") + geom_histogram()  + ggtitle("Distribution of population for US counties") + ylab("Number of counties") + xlab("Total Population")

ggplot(county_df[county_df$TotalPop < 1500000,]) + geom_histogram(aes(x = TotalPop), fill = "Gold", colour ="black") + ggtitle("Distribution of population for US counties") + ylab("Number of counties") + xlab("Total Population")
```


**Hacker Extra** - Create a function that uses the
[Freedman-Diaconis](https://en.wikipedia.org/wiki/Freedman%E2%80%93Diaconis_rule) rule for calculating bin width and number of bins. The function should take a vector (e.g. a dataframe column) as its
input and return the integer number of bins recommended for the histogram.

```{r}
numbins_FD <- function(data_col){2 * IQR(data_col) / (length(data_col))^(1/3)}

numbins_FD(county_df$TotalPop)
  

```



### Problem 4 - How do Michigan's peninsulas compare?

Create a grouped violin plot showing Income by peninsula. Make sure
your graph has a title and good axis labels.

Do the peninsulas 
differ with respect to distribution of median income over the counties?
Yes, the lower peninsula is distributed much wider. Also, the mode is much lower in the UP, roughly 10000 lower, compared to the LP. The range of the Lower Peninsula is from 30000-75000. The Upper peninsula is from 33000 to 46000. The data is fairly symmetrical for the UP and right skewed for LP.

```{r vp_inc_by_peninsula}
ggplot(data = county_MI) + geom_violin(aes(x = peninsula, y = Income)) +ggtitle("Median Income by Peninsula")+ylab("Median Income") + xlab("Peninsula")
  
```

Create several boxplots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

The lower peninsula has higher Total Population median, Q3, and max values. The spread is significantly larger for the LP, however, the minimums are pretty similar. The UP has many more Public workers than the LP, in fact the median percentage of the UP is approximately the maximum percentage of the LP. Office work is not very different in regards to median, however the LP has a little wider spread than the UP, with a higher max. Unemployment is not drastically different between peninsulas. Income per capita is much higher in the LP than the UP.

```{r boxplots}
ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = TotalPop))
ggplot(county_MI) + geom_boxplot(aes(x =peninsula, y = PublicWork)) 
ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = Office)) 
ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = Unemployment)) 
ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = Carpool)) 
ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = Poverty)) 
ggplot(county_MI) + geom_boxplot(aes(x = peninsula, y = IncomePerCap)) 

```

In county_df you'll see that there's a Region field. Create a set of faceted 
histograms showing how unemployment rate compares by region.  Make sure your graph has a title and good axis
labels. 

```{r faceted_histo}

ggplot(county_df) + geom_histogram(aes(x=Unemployment),fill = "turquoise") + xlab("Region") + ggtitle("Histogram of Unemployment Rate by Region") +facet_wrap(~Region)

```

Now do a group of boxplots of mean commute time by region.

```{r bp_commute_region}

ggplot(county_df) + geom_boxplot(aes(x="label", y=MeanCommute)) + facet_wrap(~Region) + ggtitle("Histogram of Mean Commute by Region") + ylab("Mean Commute Time")

  
  
```

### Problem 6 - Basic bar chart

Create a bar chart showing the number of counties by Division in the US. Make sure
your graph has a title and good axis labels. Figure out how to format the
x-axis labels so that they don't obscure each other.

```{r bar_counties_by_division}
ggplot(county_df) + geom_bar(aes(x = Division), color = "blue",stat = "count") + ggtitle("Counties per Division") + xlab("Division") + ylab("Number of counties") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Problem 7 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
Income and MeanCommute.

```{r xy_inc_commute}
ggplot(county_MI) + geom_point(aes(x = Income, y = MeanCommute))
```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula and the size is mapped to total population.

```{r xy_inc_commute_peninsula}
ggplot(county_MI) + geom_point(aes(x = Income, y = MeanCommute, colour = peninsula, size = TotalPop)) + ggtitle("Commute to Income") + xlab("Income") + ylab("Commute")
```

Your thoughts on the relationship between these variables?

There isn't a clear pattern, but it's easy to see that you are more likely to be in the LP if your commute is larger or your income is higher. One may be able to say that as the median income increases, the commute to work increases. This is seen as the bubbles shift up and to the right. 

Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?

The patterns we observed are pretty similar across the Bridge so perhaps the UP isn't quite as different as legend has it. 
Though we did find fairly similar results, some were different. For example, it may be easy to think that with more unemployment, you have more poverty. This is the case in the Lower Peninsula but not in the Upper. Moreover, if there is higher poverty, you might assume people to carpool more. This is seen in the Upper Peninsula but it is not as clear to see in the Lower Peninsula. The last difference we show is in the Professional vs Mean Commute graph. The interesting thing we found was how much longer the mean commute is in the Lower Peninsula vs the Upper Peninsula. The median of the Lower Peninsula is near the max of the Upper Peninsula in terms of Mean Commute.
```{r xy_faceted_peninsula}

ggplot(county_MI) + geom_point(aes(y = Income, x = Professional)) +
  facet_wrap(~peninsula)
ggplot(county_MI) + geom_point(aes(x = Professional, y = MeanCommute)) +
  facet_wrap(~peninsula)
ggplot(county_MI) + geom_point(aes(x = Poverty, y = Carpool)) +
  facet_wrap(~peninsula)
ggplot(county_MI) + geom_point(aes(x = Unemployment, y = Poverty)) +
  facet_wrap(~peninsula)

```


Now use the full US data frame.

Create a scatter plot of TotalPop by Income.

```{r xy_pop_inc}
ggplot(county_df) + geom_point(aes(x = TotalPop, y = Income)) + ggtitle("Income to Total Population") + ylab("Income") + xlab("Population")

ggplot(county_df) + geom_point(aes(x = TotalPop, y =Income)) + ggtitle("Income to Total Population") + ylab("Income") + xlab("Population") + scale_x_log10()


```

The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.

### Hacker Extra - More analysis

Do some additional plotting to look for interesting patterns or relationships
either in the full US dataset or the MI only dataset. In addition, try out
some of ggplots additional features such as themes, axis formatting, gridline
control, or others.